---
title: "Data Analysis"
author: "Cecily Zheng"
date: "`r Sys.Date()`"
output:
  pdf_document: default
  html_document:
    df_print: paged
---
```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(ggplot2)
library(dplyr)
library(tidyr)
library(stats)
library(reshape2)
library(exactRankTests)
library(kableExtra)
library(qqplotr)
```

\newpage
```{r group data, echo=FALSE}
# data
Generation <- c('F0', 'F1', 'F2', 'F3')
Light_Females_w_ <- c(10, 5, 7, 4)
Light_Males_w_ <- c(10, 16, 11, 14)
Dark_Females_w_ <- c(10, 7, 10, 7)
Dark_Males_w_ <- c(10, 14, 12, 16)
Light_Females_wt <- c(10, 15, 13, 16)
Light_Males_wt <- c(10, 4, 9, 6)
Dark_Females_wt <- c(10, 13, 10, 13)
Dark_Males_wt <- c(10, 6, 8, 4)
Frequency_Light <- c(0.50, 0.80, 0.55, 0.70)
Frequency_Dark <- c(0.50, 0.70, 0.60, 0.80)

# phenotype frequencies
df1 <- data.frame(
  Generation, 
  Light_Females_w_, 
  Light_Males_w_, 
  Dark_Females_w_, 
  Dark_Males_w_, 
  Light_Females_wt, 
  Light_Males_wt, 
  Dark_Females_wt, 
  Dark_Males_wt)

# allele frequencies
df <- data.frame(Generation, Frequency_Light, Frequency_Dark)
knitr::kable(df, booktabs = TRUE, caption = "Frequencies of White-eyed and Wild Type Alleles Across Generations")
```

```{r, class data, echo=FALSE}
# import class data for F3 generation
Fly_data <- read.csv("Fly_Data_F3_Class.csv")

# Reformat datasheet for plotting
Plotting_Data <- rbind( data.frame(Allele = "w-",
                                   Treatment = "Dark",
                                   Allele_frequency = Fly_data$Dark_Treatment_Frequency_for_white_eyed_Allele),
                        data.frame(Allele = "w-",
                                   Treatment = "Light",
                                   Allele_frequency = Fly_data$
Light_Treatment_Frequency_for_white_eyed_Allele))

# Create a variable combining species and treatment
Plotting_Data$Group <- as.factor(paste(Plotting_Data$Treatment,
                                       Plotting_Data$Allele,
                                       sep = " "))   
```

\newpage
```{r descriptive statistics group, echo=FALSE}
# group data
df_long <- melt(df, id.vars = 'Generation')
desc_stats <- df_long %>% 
  group_by(variable) %>% 
  summarise(mean = mean(value), 
            sd = sd(value),
            median = median(value),
            min = min(value), 
            max = max(value),
            iqr = IQR(value)) %>%
  ungroup() # Ensure this is called to avoid grouping in the final table output
knitr::kable(desc_stats, booktabs = TRUE, caption = "Descriptive Statistics of Group Allele Frequencies") 
```

```{r descriptive statistics class, echo=FALSE}
# class data
summary_Plotting_Data <- Plotting_Data %>%
  group_by(Group) %>%
  summarize(
    mean_allelefr = mean(Allele_frequency, na.rm = TRUE),
    sd_allelefr = sd(Allele_frequency, na.rm = TRUE),
    median_allelefr = median(Allele_frequency, na.rm = TRUE),
    min_allelefr = min(Allele_frequency, na.rm = TRUE),
    max_allelefr = max(Allele_frequency, na.rm = TRUE),
    iqr_allelefr = IQR(Allele_frequency, na.rm = TRUE),
    sample_size = n(),
  )

summary_Plotting_Data %>%
  kbl(col.names = c("Treatment", "Mean", "Standard Deviation", "Median", "Minimum", "Maximum", "Interquartile Range", "Sample Size"),
      align = 'c', caption = "Descriptive Statistics of Class F3 Allele Frequencies") %>%
  kable_classic(full_width = F, html_font = "Cambria") 
```

```{r Plot Group, echo=FALSE}
# Plotting a bar chart for allele frequency across generations
df_long_stats <- df_long %>% 
  left_join(desc_stats, by = "variable")

# Plotting a boxplot for allele frequency comparison
boxplot <- ggplot(df_long, aes(x = variable, y = value)) +
  geom_boxplot() +
  theme_minimal() +
  labs(x = 'Treatment', y = 'Frequency', 
       caption = "Figure 1: Boxplot of Comparison of White-eyed Allele Frequencies in Light and Dark Conditions") +
  theme(plot.caption = element_text(hjust = 0.5))
print(boxplot)

```

```{r, Plot Class, echo=FALSE}
# histograms used to view the distribution of data

histogram <- ggplot(Plotting_Data, aes(x = Allele_frequency, fill = Group))+
  geom_histogram(color = "#e9ecef", alpha = 0.6)+
  facet_wrap(~Group)+
  labs(x = "Average Allele Frequency",
       y = "Count",
       caption = "Figure 2: Histogram of Average White-eyed Allele Frequencies in Light and Dark Conditions") +
  theme(plot.caption = element_text(hjust = 0.5))
print(histogram)


# box plot
boxplot_class <- boxplot(Plotting_Data$Allele_frequency ~ Plotting_Data$Group,
                         xlab = "Treatment",
                         ylab = "Frequency",
                         pch = 19)

# Adding caption below the figure
mtext("Figure 3: Boxplot of Fly White-eyed Allele Frequency Across Treatments", 
      side = 1, line = 4, cex = 0.8)

```
# normality test

A normal *QQ plot* is a scatter plot created by plotted two sets of quantiles against each other. Theoretical quantiles are plotted on the x-axis and the ordered values for our variable of interest are on the y-axis, If the points plotted on the graph lie on a straight line then we can say the data distribution is normal.

*Shapiro-Wilk Normality Test* $\alpha = 0.05$, null hypothesis $H_0$: data follow normal distribution.

```{r normality test, echo=FALSE}
# Performing Shapiro-Wilk Normality Test

# group data
normality_light <- shapiro.test(df$Frequency_Light)
normality_dark <- shapiro.test(df$Frequency_Dark)
print(normality_light)
print(normality_dark)
# class data
Plotting_Data %>%
  group_by(Group) %>%
  summarize('W Stat' = shapiro.test(Allele_frequency)$statistic,
            p.value = shapiro.test(Allele_frequency)$p.value)

# Performing qqplot on Class Data
ggplot(data = Plotting_Data, mapping = aes(sample = Allele_frequency, color = Group, fill = Group))+
  stat_qq_band(alpha = 0.5, conf = 0.95, qtype = 1, bandType = "boot") +
  stat_qq_line(identity = TRUE) +
  stat_qq_point(col = "black")+
  facet_wrap(~Group, scales = "free") +
  labs(x = "Theoretical Quantiles", 
       y = "Sample Quantiles",
       caption = "Figure 4: QQ-Plot") +  theme(plot.caption = element_text(hjust = 0.5))

```

# Mann-Whitney U Test

*Mann-Whitney U Test* compare ranks (medians) between two unpaired samples. $\alpha = 0.05$, null hypothesis $H_0$: the distributions of the allele frequencies between treatments are equal.

```{r Mann-Whitney U Test, echo=FALSE}
# Performing Mann-Whitney U Test
# group data
mannwhitney_results <- wilcox.test(df$Frequency_Light, df$Frequency_Dark, na.rm = TRUE, conf.int = TRUE)
print(mannwhitney_results) 

# class data
mwu_allelefrequency <- wilcox.test(Allele_frequency ~ Group, data = Plotting_Data, na.rm = TRUE, paired = FALSE, exact = FALSE, conf.int = TRUE)
print(mwu_allelefrequency)
```

# t-test

```{r T test, echo=FALSE}
t_test_result <- t.test(df$Frequency_Light, df$Frequency_Dark, na.rm = TRUE)
print(t_test_result)
```
